---
title: "Climate Change by Team Climate.DF"
author: "Aditya Narayanan, Abhinav Tripathy"
output: html_document
---

# Introduction

Climate.DF is a project created by data-oriented student Climatologists who seek to keep the world aware of how climate has changed from the past and how it is going to change in the future. We intend to do it using statistical methods, data visualization and machine learning models.

### Importance of Climate Change

Understanding and predicting what the future weather might bring, or predicting how climate will change over the next few decades is of vital importance - both for our economy and for society.
Climate can be thought of as the average or typical weather conditions we experience. Scientists know that climate varies naturally on many timescales and they know that people are affecting climate - particularly through emissions of greenhouse gases.

### What do we wish to achieve?

We are focusing on different accurate and comprehendible visualizations and portraying how factors like temperature, carbon dioxide, and different natural gases disrupt climate patterns. Through regression modeling and time series analysis, we aim to forecast future patterns of climate and provide methods to control climate change using statistical methods. 
Our project showcases prehistoric climate changes in an interactive Shiny Application where the user can select a start and end data and view different temperature changes over the years. We are interested in showcasing different statistics to portray how climate has changed over different geographies. With respect to time-series analysis, we aim to do a global forecast of temperature to give our viewers an idea of the different global temperature changes and use ARIMA models to forecast future temperature changes. Using effective visualizations of maps, we aim to show how different anomalies like temperatures and precipitation change with different continents or countries.

The main goal of this project is to showcase the reasons for rise in temperatures and how the temperatures are rising across the world. There are two main datasets we will be focusing on, the first one is data collected from Berkeley Earth Temperature Data, and the second being a dataset from Kaggle which contains information about temperatures and greenhouse gases.

## Part 1 
##### Exploratory Data Analysis and Modeling of historical data from 1983-2008
In this part of the project, we will be analyzing how climate in the past has changed due to the effect of greenhouse gases.


```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(plotly)
library("RColorBrewer")
library(lubridate)
library(ggpubr)
library(TTR)
library(ggridges)
library(forecast)
library(rWBclimate)
library(ggmap)
library(maps)
library(maptools)
library(choroplethr)
library(choroplethrMaps)
library(pander)
options(warn=-1)
```


```{r, echo=FALSE}
climate <- read.csv("../Datasets/climate_change.csv")

if(nrow(climate) != nrow(na.omit(climate))) {
  climate <- na.omit(climate)
}
```
We have successfully loaded our data and handled missing values.

In this dataset, all of the months are given in numbers from 1-10, so in order to increase clarity of the dataset, the numbers have been changed to months.

```{r, echo=FALSE}
climate$Month[climate$Month == 1]<- "January"
climate$Month[climate$Month == 2]<- "February"
climate$Month[climate$Month == 3]<- "March"
climate$Month[climate$Month == 4]<- "April"
climate$Month[climate$Month == 5]<- "May"
climate$Month[climate$Month == 6]<- "June"
climate$Month[climate$Month == 7]<- "July"
climate$Month[climate$Month == 8]<- "August"
climate$Month[climate$Month == 9]<- "September"
climate$Month[climate$Month == 10]<- "October"
climate$Month[climate$Month == 11]<- "November"
climate$Month[climate$Month == 12]<- "December"
```

---

#### Data Exploration

Greenhouse gases are poisonous to our atmosphere and causes various changes to atmospheric temperatures. Through a bar chart, let's explore how these greenhouse gases have caused a change in temperature in the past.

Note: Hover over the barplots to view interact with graph. 

```{r, echo=FALSE}
climate.plot.CO2 <- ggplot(climate, aes(x = Year, y = Temp, fill = CO2)) + xlab("Year") +ylab("Temperature")
plot.CO2 <- climate.plot.CO2 + geom_bar( position = "dodge", stat = "identity",color= "white")
ggplotly(plot.CO2)

climate.plot.CH4 <- ggplot(climate, aes(x = Year, y = Temp, fill = CH4)) + xlab("Year") +ylab("Temperature") 
plot.CH4 <- climate.plot.CH4 + geom_bar( position = "dodge", stat = "identity",color= "white")
ggplotly(plot.CH4)

climate.plot.CFC11 <- ggplot(climate, aes(x = Year, y = Temp, fill = CFC.11)) + xlab("Year") +ylab("Temperature")
plot.CFC11 <- climate.plot.CFC11 + geom_bar( position = "dodge", stat = "identity",color= "white")
ggplotly(plot.CFC11)

climate.plot.CFC12 <- ggplot(climate, aes(x = Year, y = Temp, fill = CFC.12)) + xlab("Year") +ylab("Temperature") 
plot.CFC12 <- climate.plot.CFC12 + geom_bar( position = "dodge", stat = "identity",color= "white")
ggplotly(plot.CFC12)
```

* As we can see, over the years, the Carbon Dioxide (CO2) and Methane (CH4) levels started off very low, but as time proceeds, we can see the levels of these gases grow rapidly causing larger changes to average temperatures. These changes of CO2 and CH4 levels suggests advancements in industrial districts or factories around the world, which emit a lot of greenhouse gases.  

* As the CFCs increase, they cause damage to our stratospheric ozone layer hence allowing more UV rays to enter the Earth and increasing temperatures worldwide. CFCs are used primarily as a foaming agent for building insulation, refrigerators and other consumer products, and countries with mass industrial development such as China, USA, UK, India, and Japan are primary emmiters for this chemical. 

---


On camparison, we would like to portray how temperature changes with important factors such as CO2, CH4, N2O, and MEI.

```{r, echo=FALSE}
scat_co2 <- ggplot(climate, aes(Temp, CO2))+geom_point(color="blueviolet")+geom_smooth(method = "lm")+ggtitle("CO2 vs Temp")

scat_ch4 <- ggplot(climate, aes(Temp, CH4))+geom_point(color="springgreen4")+geom_smooth(method = "lm")+ggtitle("CH4 vs Temp")

scat_n2o <- ggplot(climate, aes(Temp, N2O))+geom_point(color="orange")+geom_smooth(method = "lm")+ggtitle("N2O vs Temp")

scat_MEI <- ggplot(climate, aes(Temp, MEI))+geom_point(color="darkred")+geom_smooth(method = "lm")+ggtitle("MEI vs Temp")

gas_scatter<-ggarrange(scat_co2, scat_ch4, scat_n2o, scat_MEI + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
annotate_figure(gas_scatter,
                top = text_grob("Vartations of CO_2, N2O, CH4 and MEI by Temperature", color = "black", face = "bold", size = 13))

```

---

Now, we will focus on a small time series analysis to visualize how temperature, CO2, CH4, N20 and MEI have changed in the past, over time.

(a) Temperature
```{r, echo=FALSE}
climate.dates <- climate %>%
  mutate(date = ymd(paste(climate$Year, climate$Month, truncated = 1)))

climate.vec.temp <- climate.dates$Temp
climate.temp.ts <- ts(climate.vec.temp, frequency = 12, start = c(1983,1))
plot.temp.ts <- plot.ts(climate.temp.ts, ylab="Temperature Change", col=1)

```

Through this time series visualization, we can observe that there is an upward trend in the increase in temperature from 1983-2008.

```{r, echo=FALSE}
temp.hw <- HoltWinters(climate.temp.ts)
temp.pred.hw <- forecast:::forecast.HoltWinters(temp.hw, h=20)
forecast:::plot.forecast(temp.pred.hw)
```
Using Holt-Winter's statistical time series predictions, we can observe how the temperatures are bound to change in the forthcoming years. 

(b) Carbon Dioxide (CO2)
```{r, echo=FALSE}
climate.vec.co2 <- climate.dates$CO2
climate.co2.ts <- ts(climate.vec.co2, frequency = 12, start = c(1983,1))
plot.co2.ts <- plot.ts(climate.co2.ts, ylab="Carbon Dioxide Change", col=2)

```

* Extra carbon dioxide in the atmosphere increases the greenhouse effect. Thermal energy is trapped by the atmosphere, causing the planet to become warmer than it would be naturally. This increase in the Earth's temperature is called global warming.
* From the graph it is clear that CO2 levels have been constantly increasing in atmosphere.

```{r, echo=FALSE}
co2.hw <- HoltWinters(climate.co2.ts)
co2.pred.hw <- forecast:::forecast.HoltWinters(co2.hw, h=20)
forecast:::plot.forecast(co2.pred.hw)
```

Using Holt-Winter's statistical time series predictions, we can observe how the CO@ levels are bound to change in the forthcoming years. The increase in CO2 levels seem to be linearly increasing as time proceeds. 

(c) Methane (CH4)
```{r, echo=FALSE}
climate.vec.ch4 <- climate.dates$CH4
climate.ch4.ts <- ts(climate.vec.ch4, frequency = 12, start = c(1983,1))
plot.ch4.ts <- plot.ts(climate.ch4.ts, ylab="Methane Change", col=3)

```

* Methane bubbles are effect and cause of rise in temperature. Due to climate change, more methane is bubbling up from water bodies throughout the world. The release of methane, a potent greenhouse gas, leads to a further increase in temperature.
* If methane leaks into the air before being used it absorbs the sun's heat, warming the atmosphere. For this reason, it's considered a greenhouse gas, like carbon dioxide.
* From the graph, we can see that there has been an increase in methane over time.

```{r, echo=FALSE}
ch4.hw <- HoltWinters(climate.ch4.ts)
ch4.pred.hw <- forecast:::forecast.HoltWinters(ch4.hw, h=20)
forecast:::plot.forecast(ch4.pred.hw)
```

Using Holt-Winter's statistical time series predictions, we can observe how the CH4 are bound to change in the forthcoming years. The increase in CH4 levels seem to be following a logarithmic pattern as time proceeds. The more precautions we take, the lower we will find our CH4 levels over time and therefore, temperatures will in turn decrease. 


(d) Nitrous Oxide (N2O)
```{r, echo=FALSE}
climate.vec.n2o <- climate.dates$N2O
climate.n2o.ts <- ts(climate.vec.n2o, frequency = 12, start = c(1983,1))
plot.n2o.ts <- plot.ts(climate.n2o.ts, ylab="Nitrous Oxide Change", col=4)

```

* Nitrous Oxide has shown a sharp increase over the years which is very harmful for the atmosphere, and it is more harmful than CO2 and CH4 on the global warming index, and we must prevent the expulsion of this gas. 

```{r, echo=FALSE}
n2o.hw <- HoltWinters(climate.n2o.ts)
n2o.pred.hw <- forecast:::forecast.HoltWinters(n2o.hw, h=20)
forecast:::plot.forecast(n2o.pred.hw)
```

Using Holt-Winter's statistical time series predictions, we can observe how N2O levels are bound to change in the forthcoming years. The increase in N2O levels seem to be sharply increasing as time proceeds. 


(e) Multivariate EI Nino Southern Oscillation Index (MEI)
```{r, echo=FALSE}
climate.vec.MEI <- climate.dates$MEI
climate.MEI.ts <- ts(climate.vec.MEI, frequency = 12, start = c(1983,1))
plot.MEI.ts <- plot.ts(climate.MEI.ts, ylab="MEI Change", col=2)

```

* There is a seasonality to MEI but not a continuous positive or negative trend over the years.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(climate, aes(x = Temp, y = as.factor(Year), fill=stat(x))) + 
  geom_density_ridges_gradient(scale = 3, size = 0.3, alpha = 0.5) + scale_fill_viridis_c(
                       name = "Temp[C]") + labs(title='Temperature Density Function Plot ')+ylab('Year')+xlab('Temperature')
```

Over the years, there has been an increase in temperature, but there has been a slight decrease from 2005-2008. The average increase in temperatures is still greater after 2000. 

---

#### Modeling

We will be constructing linear regression models to show how temperature is affected by many factors.

```{r, echo=FALSE}
# Create the model
fit.temp.co2 <- lm(Temp ~ CO2, data = climate)
fit.temp.co2$predicted <- predict(fit.temp.co2)
reg.plot.co2 <- ggplot(climate, aes(x = CO2, y = Temp)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Plot regression slope
  geom_segment(aes(xend = CO2, yend = fit.temp.co2$predicted), alpha = .2) +  # alpha to fade lines
  geom_point(color="blue") +
  geom_point(aes(y = fit.temp.co2$predicted), color='darkgreen', shape = 1) +
  theme_bw() 

fit.temp.ch4 <- lm(Temp ~ CH4, data = climate)
fit.temp.ch4$predicted <- predict(fit.temp.ch4)
reg.plot.ch4 <- ggplot(climate, aes(x = CH4, y = Temp)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Plot regression slope
  geom_segment(aes(xend = CH4, yend = fit.temp.ch4$predicted), alpha = .2) +  # alpha to fade lines
  geom_point(color="blue") +
  geom_point(aes(y = fit.temp.ch4$predicted), color='darkgreen', shape = 1) +
  theme_bw() 

fit.temp.n2o <- lm(Temp ~ N2O, data = climate)
fit.temp.n2o$predicted <- predict(fit.temp.n2o)
reg.plot.n2o <- ggplot(climate, aes(x = N2O, y = Temp)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Plot regression slope
  geom_segment(aes(xend = N2O, yend = fit.temp.n2o$predicted), alpha = .2) +  # alpha to fade lines
  geom_point(color="blue") +
  geom_point(aes(y = fit.temp.n2o$predicted), color='darkgreen', shape = 1) +
  theme_bw()

fit.temp.cfc11 <- lm(Temp ~ CFC.11, data = climate)
fit.temp.cfc11$predicted <- predict(fit.temp.cfc11)
reg.plot.cfc11 <- ggplot(climate, aes(x = CFC.11, y = Temp)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Plot regression slope
  geom_segment(aes(xend = CFC.11, yend = fit.temp.cfc11$predicted), alpha = .2) +  # alpha to fade lines
  geom_point(color="blue") +
  geom_point(aes(y = fit.temp.cfc11$predicted), color='darkgreen', shape = 1) +
  theme_bw()

fit.temp.cfc12 <- lm(Temp ~ CFC.12, data = climate)
fit.temp.cfc12$predicted <- predict(fit.temp.cfc12)
reg.plot.cfc12 <- ggplot(climate, aes(x = CFC.12, y = Temp)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Plot regression slope
  geom_segment(aes(xend = CFC.12, yend = fit.temp.cfc12$predicted), alpha = .2) +  # alpha to fade lines
  geom_point(color="blue") +
  geom_point(aes(y = fit.temp.cfc12$predicted), color='darkgreen', shape = 1) +
  theme_bw()

gas_reg <- ggarrange(reg.plot.co2, reg.plot.ch4, reg.plot.n2o, reg.plot.cfc11, reg.plot.cfc12 + rremove("x.text"), 
          labels = c("A", "B", "C", "D", "E"),
          ncol = 3, nrow = 2)
annotate_figure(gas_reg,
                top = text_grob("Regression Analysis", color = "black", face = "bold", size = 13))
```


In this interactive plot, we can see our actual values plotted in blue, and our predicted values plotted in dark green. The predicted values lie on our regression lines, proving that these models aew very efficient, and that indeed, Temperature rises with increase in greenhouse gas emission levels.

--- 

#### Showcasing Monthly Temperature Changes Group by Year. 
###### Animations of the Visualizations

```{r, echo=FALSE}
library(gganimate)
plot.animation <- ggplot(climate, aes(x = Year, y = Temp, color = Month, group = Month)) +
    geom_path() +
    geom_point() +
    facet_wrap(~ Month) +
    theme(legend.position = 'none') +
    labs(title = 'Temperature Variation, Year: {frame_along}') +
    transition_reveal(along = Year) +
    ease_aes('linear')
animate(plot.animation, 100, 10, renderer = gifski_renderer("tempanim.gif"))
```

We can observe that this animation showcases the change in temperature per month for every year included in our past data. 

---

# Part 2
**In this part we will be changin our dataset to more recent Global Temperature data acquired from Berkeley Earth Temperature Data and the World Bank Climate Repository**

### Interactive Shiny Application for Temperature Changes from 1985-2015

The user may interact with this SHiny application to allow them to visualize changes in temperatures slelecting their own start and ending year. The user may select to visualize either just Land Temperature data or even the combined Land + Ocean temperature data. The user also has the option of viewing the best fit line through the line graph. 

Please click on the link below to view the Shiny App:

<font size="3"><strong>[Global Temperature Changes](https://adityanar.shinyapps.io/dashboard1/)</strong></font>

---

### Shiny Application of Temperature Changes Around the World

A Shiny Application has been created to visualize changes in temperature around the world. To visualize changes, please click on the link below to view the Shiny App:

<font size="3"><strong>[Temperature Changes on a Map](https://adityanar.shinyapps.io/Maps/)</strong></font>

--- 

### Land Temperature Changes in the United States

```{r, echo=FALSE, message=FALSE, warning=FALSE}
temp_state <- read.csv("../Datasets/GlobalLandTemperaturesByState.csv")
temp_usa <- temp_state %>% filter(Country=="United States") %>% separate(col = dt, into = c("Year", "Month", "Day"), convert = TRUE) -> temp_usa
temp_usa<-na.omit(temp_usa)
temp_usa %>%
  filter(State!="Hawaii" & State!="Alaska") -> temp_usa1
temp_usa1 <- na.omit(temp_usa1)

temp_usa1 %>% filter(Year>1850) %>% group_by(Year) %>% summarise(Temp=mean(AverageTemperature)) -> temp_usa2
```

Now, we will create a plot of the Average Temperature of the US from 1850 onwards.

```{r, echo=FALSE}
qplot(Year, Temp, data=temp_usa2, main="US Average Temperature From 1850 Onwards",geom=c("point","smooth"))+ aes(colour = Temp) + scale_color_gradient(low="blue", high="red")

```

Looks like the average temperature has risen over time, from this plot.

Using an analysis of variance test, we can check if there are any differences in Average Temperature in the US. 
```{r, echo=FALSE}
temp_usa1 %>% filter(Year==1850 | Year==1890 | Year==1930 | Year==1970 | Year==2013) %>%
  group_by(State,Year) %>% 
  summarise(Temp = mean(AverageTemperature)) -> temp_usa3
temp_usa3$Year <- as.factor(temp_usa3$Year)
aovtest <- (aov(Temp~Year, data=temp_usa3)) 
summary(aovtest) %>% pander
```

```{r, echo=FALSE}
TukeyHSD(aovtest)
```

The p-values highlight differences in the Average Temperatures at the 10% level between the years 1970 vs 2013, 1980vs 2013 and 1850 vs 2013. However, if the end year was chosen to 2010 or 2011 this effect is likely to disapear.

Finally, we will use a chloropeth map to display temperature changes by state in the US. 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
temp_usa$State <- as.character(temp_usa$State)
temp_usa$State[temp_usa$State=="Georgia (State)"] <- "Georgia"
temp_usa$State<- as.factor(temp_usa$State)                    

#' select columns of interest
temp_usa %>% 
  select(Year,AverageTemperature,State) %>%
  group_by(Year,State) %>%
  summarise(value=mean(AverageTemperature)) -> temp_usa4

#Data frame must have a column named region (all lower case) and another one value.
colnames(temp_usa4)[2]<- "region"
temp_usa4$region<-tolower(temp_usa4$region)

temp_usa4 %>%
  filter(Year==1850) -> temp_usa1850
temp_usa1850<-temp_usa1850[,2:3]

temp_usa4 %>%
  filter(Year==2013) -> temp_usa2013
temp_usa2013<-temp_usa2013[,2:3]


print(state_choropleth(temp_usa1850,
                       title="Land Temperature 1850", 
                       num_colors = 8,
                       legend="Degrees"),reference_map=TRUE)
```

```{r, echo=FALSE}
print(state_choropleth(temp_usa2013,
                       title="Land Temperature 1850", 
                       num_colors = 8,
                       legend="Degrees"),reference_map=TRUE)
```

--- 
Now we will construct a time-series analysis of our data to get a better sense on how our average temperature is going to change over the next 10 years.

```{r, echo=FALSE}
library(fpp2)
global_temp <- read.csv("../Datasets/GlobalTemperatures.csv")
global_temp$dt<-as.Date(global_temp$dt,format="yyyy-mm-dd")
options(repr.plot.width=4, repr.plot.height=4)
ggplot(global_temp,aes(x=LandAverageTemperature))+geom_histogram(aes(y=..density..),fill="blue")+geom_density(col="red")
```


We have to prepare our data for Time Series Analysis and Forecasting
```{r, echo=FALSE}
options(repr.plot.width=5, repr.plot.height=5)
temp_ts<-ts(global_temp[,2],start=c(1980,1),end=c(2015,12),frequency=12)
autoplot(temp_ts)
ggseasonplot(temp_ts)
```

The seasonal plot clearly shows that Average Temperature rises slowly from January and attains peak in June and July, and then its starts to fall again. This repeated almost every year.

```{r, echo=FALSE}
ggsubseriesplot(temp_ts)
gglagplot(temp_ts)
```

A ggsubseriesplot is used to plot trend for every month seperately, within the same plot.

--- 
**Naive Forecasting**
Forecating temperatures for the next 10 years. 
```{r, echo=FALSE}
temp_fc<-snaive(temp_ts,h=120)
autoplot(temp_fc)
summary(temp_fc) %>% pander
```

```{r, echo=FALSE}
checkresiduals(temp_fc)
```

On plotting our residuals, we can see that they are correlated. The residuals have a zero mean which shows that the forecast is unbiased. Hence, we have a good forecasting method in place. 

--- 

**ARIMA Models and Forecasting**
```{r, echo=FALSE}
fit.arima<-auto.arima(temp_ts,lambda=0)
d<-1
D<-1
fit.arima %>%forecast(h=120)%>%autoplot()
checkresiduals(fit.arima)
```

We achieve a p-value less than 0.05 which fits our 95% threshold, and this shows that our model is efficient. Again, On plotting our residuals, we can see that they are correlated. The residuals have a zero mean which shows that the forecast is unbiased. Hence, the ARIMA model has a good predictive power. 

--- 

# Conclusion
We have written an indepth analysis on what the future of climate change looks like. 

In Part 1, we analyzed and made decisions on what the future of the rises in temperature of this world would look like with the increase in greenhouse gases. We as citizens of the world have to do our best to promote a healthy environment and prevent global warming of our Earth. Using different visualizations we were able to showcase how climate change is real and is happening a rapid rate. We used Holt-Winters model to predict temperature changes given the current circumsances, using past/historic data. 

In Part 2, we made different interactive applications to visiualize how temperatures have changed thorughout the world both in land temperatures, ocean temperatures, and have geographically represented our results with maps and graphs. We used more efficiecnt Time Series models such as Naive Forecasting and ARIMA models. We were able to predict rise in global temperatures across the world for the next 10 years. 

We hope our analytics and results keep people aware of our ever changing world and how climate plays a big role in this change. If everyone takes care of their environment, we can get a cleaner and safer environment which could stop the hazardous effect of climate change on the ecosystem. 
